Using generalized additive models to analyze biomedical non-linear longitudinal data

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/01-Background.Rmd
Need to check references. refs1 works but refs does not.

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001;@roth2017], and cell metabolism [@jones2018;@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].

Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004;@schober2018].

The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006]. This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011;@skala2010;@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

@@

Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010;@halsey2015], consequently biasing the conclusions of the study. Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012;@albers2019]. Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].  Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976;@greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001]. In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.

@@

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference. In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.   Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012;@pedersen2019;@simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998] . There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically  $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, $\textsf{R}$ [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016;@wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.

>>>>

Longitudinal studies are designed to repeatedly measure a variable of interest in a group (or groups) of subjects, with the intention of observing the evolution of effect across time rather than analyzing a single time point (e.g., a cross-sectional study). Biomedical research frequently uses longitudinal studies to analyze the evolution of a "treatment" effect across multiple time points; and in such studies the subjects of analysis range from animals (mice, rats, rabbits), to human patients, cells, or blood samples, among many others. Tumor response [@roblyer2011;@tank2020;@pavlov2018;@demidov2018], antibody expression [@ritter2001] [@roth2017], and cell metabolism [@jones2018] [@skala2010] are examples of the different situations where researchers have used longitudinal designs to study some  physiological response. Because the frequency of the measurements in a longitudinal study is dependent on the biological phenomena of interest and the experimental design of the study, the frequency of such measurements can range from minute intervals to study a short-term response such as anesthesia effects in animals[@greening2018], to weekly measurements to analyze a mid-term response like the evolution of dermatitis symptoms in breast cancer patients [@sio2016], to monthly measurements to study a long-term response such as mouth opening following radiotherapy (RT) in neck cancer patients [@kamstra2015].

Traditionally, a “frequentist” or "classical" statistical paradigm is used in biomedical research to derive inferences from a longitudinal study. The frequentist paradigm regards probability as the limit of the expected outcome when an experiment is repeated a large number of times [@wagenmakers2008], and such view is applied to the analysis of longitudinal data by assuming a null hypothesis under a statistical model that is often an _analysis of variance over repeated measures_ (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model makes three key assumptions regarding longitudinal data: 1) linearity of the response across time, 2) constant correlation across same-subject measurements, and 3) observations from each subject are obtained at all time points through the study (a condition also known as _complete observations_) [@gueorguieva2004] [@schober2018].

The expected linear behavior of the response through time is a key requisite in rm-ANOVA [@pinheiro2006]. This "linearity assumption" in rm-ANOVA implies that the model is misspecified when the data does not follow a linear trend, which results in unreliable inference. In biomedical research, non-linear trends are the norm rather than the exception in longitudinal studies. A particular example of this non-linear behavior in longitudinal data arises in measurements of tumor response to chemo and/or radiotherapy in preclinical and clinical settings [@roblyer2011] [@skala2010] [@vishwanath2009]. These studies have shown that the collected signal does not follow a linear trend over time, and presents extreme variability at different time points, making the fit of rm-ANOVA model inconsistent with the observed variation. Therefore, when rm-ANOVA is used to draw inference of such data the estimates are inevitably biased, because the model is only able to accommodate linear trends that fail to adequately represent the biological phenomenon of interest.

@@

Additionally, the _p-value_ itself is highly variable, and multiple comparisons can inflate the false positivity rate (Type I error or $\alpha$) [@liu2010; @halsey2015], consequently biasing the conclusions of the study. Corrections exist to address the Type I error issue of multiple comparisons (such as Bonferroni [@abdi2010]), but they in turn reduce statistical power (1-$\beta$)[@nakagawa2004], and lead to increased Type II error (failing to reject the null hypothesis when the null hypothesis is false) [@gelman2012; @albers2019]. Therefore, the tradeoff of _post hoc_ comparisons in rm-ANOVA between Type I, II and III errors might be difficult to resolve in a biomedical longitudinal study where a delicate balance exists between statistical power and sample size.

On the other hand, the assumption of constant correlation in rm-ANOVA (often known as the _compound symmetry assumption_) is typically unreasonable because correlation between the measured responses often diminishes as the time interval between the observation increases [@ugrinowitsch2004].  Corrections can be made in rm-ANOVA in the absence of compound symmetry [@huynh1976; @greenhouse1959], but the effectiveness of the correction is limited by the size of the sample, the number of measurements[@haverkamp2017], and group sizes [@keselman2001]. In the case of biomedical research, where living subjects are frequently used, sample sizes are often not "large" due to ethical and budgetary reasons [@charan2013] which might cause the corrections for lack of compound symmetry to be ineffective.

@@

As the rm-ANOVA and the more flexible LMEM approaches make overly restrictive assumptions regarding the linearity of the response, there is a need for biomedical researchers to explore the use of additional statistical tools that allow the data (and not an assumption in trend) to determine the trend of the fitted model, to enable appropriate inference. In this regard, generalized additive models (GAMs) present an alternative approach to analyze longitudinal data. Although not frequently used by the biomedical community, these semi-parametric models are customarily used in other fields to analyze longitudinal data.   Examples of the use of GAMs include the analysis of temporal variations in geochemical and palaeoecological data  [@rose2012; @pedersen2019; @simpson2018], health-environment interactions [@yang2012] and the dynamics of government in political science [@beck1998]. There are several advantages of GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of smooth responses that enable the data to dictate the trend in the fit of the model, 2) they can model non-constant correlation between repeated measurements [@wood2017] and 3) can easily accommodate missing observations. Therefore, GAMs can provide a more flexible statistical approach to analyze non-linear biomedical longitudinal data than LMEMs and rm-ANOVA.

The current advances in programming languages designed for statistical analysis (specifically  $\textsf{R}$), have eased the computational implementation of traditional models such as rm-ANOVA and more complex approaches such as LMEMs and GAMs. In particular, $\textsf{R}$ [@r] has an extensive collection of documentation and functions to fit GAMs in the package _mgcv_ [@wood2016; @wood2017] that not only speed up the initial stages of the analysis but also enable the use of advanced modeling structures (e.g. hierarchical models, confidence interval comparisons) without requiring advanced programming skills from the user. At the same time, $\textsf{R}$ has many tools that simplify data simulation, an emerging strategy used to test statistical models [@haverkamp2017]. Data simulation methods allow the researcher to create and  explore different alternatives for analysis without collecting information in the field, reducing the time window between experiment design and its implementation, and simulation can be also used for power calculations and study design questions.

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.Rmd
Need to check references. refs1 works but refs does not.

---
title: Using generalized additive models to analyze biomedical non-linear longitudinal data
runninghead: Mundo \emph{et al}.
author:
- name: Ariel I. Mundo
  num: 1
- name: John R. Tipton
  num: 2
- name: Timothy J. Muldoon*
  num: 1
address:
- num: 1
  org: Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US
- num: 2
  org:  Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
corrauth: "Corresponding author"
email: tmuldoon@uark.edu
abstract: "In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research. In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
keywords: Cancer biology; Tumor response; generalized additive models; simulation; R
classoption:
  - Royal
  - times
bibliography: refs1.bib
bibliographystyle: sagev
output:
  rticles::sage_article:
    keep_tex: yes
    #pandoc_args: --listings
    fig_caption: yes
    #extra_dependencies:
      #subfig: null #allows for subfigures
      #breqn: null #line breaks for long equations
      #caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
---

>>>>

---
title: Using generalized additive models to analyze biomedical non-linear longitudinal data
runninghead: Mundo \emph{et al}.
author:
- name: Ariel I. Mundo
  num: 1
- name: John R. Tipton
  num: 2
- name: Timothy J. Muldoon*
  num: 1
address:
- num: 1
  org: Department of Biomedical Engineering, University of Arkansas, Fayetteville, AR, US
- num: 2
  org:  Department of Mathematical Sciences, University of Arkansas, Fayetteville, AR, US
corrauth: "Timothy J. Muldoon"
email: tmuldoon@uark.edu
abstract: "In biomedical research, the outcome of longitudinal studies has been traditionally analyzed using the _repeated measures analysis of variance_ (rm-ANOVA) or more recently, _linear mixed models_ (LMEMs). Although LMEMs are less restrictive than rm-ANOVA in terms of correlation and missing observations, both methodologies share an assumption of linearity in the measured response, which results in biased estimates and unreliable inference when they are used to analyze data where the trends are non-linear, which is a common occurrence in biomedical research. In contrast, _generalized additive models_ (GAMs) relax the linearity assumption, and allow the data to determine the fit of the model while permitting missing observations and different correlation structures. Therefore, GAMs present an excellent choice to analyze non-linear longitudinal data in the context of biomedical research. This paper summarizes the limitations of rm-ANOVA and LMEMs and uses simulated data to visually show how both methods produce biased estimates when used on non-linear data. We also present the basic theory of GAMs, and using trends of oxygen saturation in tumors reported in the biomedical literature, we simulate example longitudinal data (2 treatment groups, 10 subjects per group, 5 repeated measures for each group) to demonstrate how these models can be computationally implemented. We show that GAMs are able to produce estimates that are consistent with the trends of biomedical non-linear data even in the case when missing observations exist (with 40% of the simulated observations missing), allowing reliable inference from the data. To make this work reproducible, the code and data used in this paper are available at: https://github.com/aimundo/GAMs-biomedical-research."
keywords: Cancer biology; Tumor response; generalized additive models; simulation; R
classoption:
  - Royal
  - times
bibliography: refs.bib
bibliographystyle: sagev
output:
  rticles::sage_article:
    keep_tex: yes
    #pandoc_args: --listings
    fig_caption: yes
    #extra_dependencies:
      #subfig: null #allows for subfigures
      #breqn: null #line breaks for long equations
      #caption: ["font={small}"] #size of the figure captions
      #float: null #allows for control of placement of figures
---

### 2021-06-18 Manuscripts/Manuscript by chapters -for submission/Full_document_SAGE.tex
Need to check references. refs1 works but refs does not.

\corrauth{Corresponding author}

@@

Longitudinal studies are designed to repeatedly measure a variable of
interest in a group (or groups) of subjects, with the intention of
observing the evolution of effect across time rather than analyzing a
single time point (e.g., a cross-sectional study). Biomedical research
frequently uses longitudinal studies to analyze the evolution of a
``treatment'' effect across multiple time points; and in such studies
the subjects of analysis range from animals (mice, rats, rabbits), to
human patients, cells, or blood samples, among many others. Tumor
response \citep{roblyer2011, tank2020, pavlov2018, demidov2018},
antibody expression \citep{ritter2001, roth2017}, and cell metabolism
\citep{jones2018, skala2010} are examples of the different situations
where researchers have used longitudinal designs to study some
physiological response. Because the frequency of the measurements in a
longitudinal study is dependent on the biological phenomena of interest
and the experimental design of the study, the frequency of such
measurements can range from minute intervals to study a short-term
response such as anesthesia effects in animals\citep{greening2018}, to
weekly measurements to analyze a mid-term response like the evolution of
dermatitis symptoms in breast cancer patients \citep{sio2016}, to
monthly measurements to study a long-term response such as mouth opening
following radiotherapy (RT) in neck cancer patients \citep{kamstra2015}.

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm
is used in biomedical research to derive inferences from a longitudinal
study. The frequentist paradigm regards probability as the limit of the
expected outcome when an experiment is repeated a large number of times
\citep{wagenmakers2008}, and such view is applied to the analysis of
longitudinal data by assuming a null hypothesis under a statistical
model that is often an \emph{analysis of variance over repeated
measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model
makes three key assumptions regarding longitudinal data: 1) linearity of
the response across time, 2) constant correlation across same-subject
measurements, and 3) observations from each subject are obtained at all
time points through the study (a condition also known as \emph{complete
observations}) \citep{gueorguieva2004, schober2018}.

The expected linear behavior of the response through time is a key
requisite in rm-ANOVA \citep{pinheiro2006}. This ``linearity
assumption'' in rm-ANOVA implies that the model is misspecified when the
data does not follow a linear trend, which results in unreliable
inference. In biomedical research, non-linear trends are the norm rather
than the exception in longitudinal studies. A particular example of this
non-linear behavior in longitudinal data arises in measurements of tumor
response to chemo and/or radiotherapy in preclinical and clinical
settings \citep{roblyer2011, skala2010, vishwanath2009}. These studies
have shown that the collected signal does not follow a linear trend over
time, and presents extreme variability at different time points, making
the fit of rm-ANOVA model inconsistent with the observed variation.
Therefore, when rm-ANOVA is used to draw inference of such data the
estimates are inevitably biased, because the model is only able to
accommodate linear trends that fail to adequately represent the
biological phenomenon of interest.

@@

As the rm-ANOVA and the more flexible LMEM approaches make overly
restrictive assumptions regarding the linearity of the response, there
is a need for biomedical researchers to explore the use of additional
statistical tools that allow the data (and not an assumption in trend)
to determine the trend of the fitted model, to enable appropriate
inference. In this regard, generalized additive models (GAMs) present an
alternative approach to analyze longitudinal data. Although not
frequently used by the biomedical community, these semi-parametric
models are customarily used in other fields to analyze longitudinal
data. Examples of the use of GAMs include the analysis of temporal
variations in geochemical and palaeoecological data
\citep{rose2012, pedersen2019, simpson2018}, health-environment
interactions \citep{yang2012} and the dynamics of government in
political science \citep{beck1998} . There are several advantages of
GAMs over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible
class of smooth responses that enable the data to dictate the trend in
the fit of the model, 2) they can model non-constant correlation between
repeated measurements \citep{wood2017} and 3) can easily accommodate
missing observations. Therefore, GAMs can provide a more flexible
statistical approach to analyze non-linear biomedical longitudinal data
than LMEMs and rm-ANOVA.

@@

\bibliographystyle{sagev}
\bibliography{refs1.bib}

>>>>

\corrauth{Timothy J. Muldoon}

@@

Longitudinal studies are designed to repeatedly measure a variable of
interest in a group (or groups) of subjects, with the intention of
observing the evolution of effect across time rather than analyzing a
single time point (e.g., a cross-sectional study). Biomedical research
frequently uses longitudinal studies to analyze the evolution of a
``treatment'' effect across multiple time points; and in such studies
the subjects of analysis range from animals (mice, rats, rabbits), to
human patients, cells, or blood samples, among many others. Tumor
response \citep{roblyer2011, tank2020, pavlov2018, demidov2018},
antibody expression \citep{ritter2001} \citep{roth2017}, and cell
metabolism \citep{jones2018} \citep{skala2010} are examples of the
different situations where researchers have used longitudinal designs to
study some physiological response. Because the frequency of the
measurements in a longitudinal study is dependent on the biological
phenomena of interest and the experimental design of the study, the
frequency of such measurements can range from minute intervals to study
a short-term response such as anesthesia effects in
animals\citep{greening2018}, to weekly measurements to analyze a
mid-term response like the evolution of dermatitis symptoms in breast
cancer patients \citep{sio2016}, to monthly measurements to study a
long-term response such as mouth opening following radiotherapy (RT) in
neck cancer patients \citep{kamstra2015}.

Traditionally, a ``frequentist'' or ``classical'' statistical paradigm
is used in biomedical research to derive inferences from a longitudinal
study. The frequentist paradigm regards probability as the limit of the
expected outcome when an experiment is repeated a large number of times
\citep{wagenmakers2008}, and such view is applied to the analysis of
longitudinal data by assuming a null hypothesis under a statistical
model that is often an \emph{analysis of variance over repeated
measures} (repeated measures ANOVA or rm-ANOVA). The rm-ANOVA model
makes three key assumptions regarding longitudinal data: 1) linearity of
the response across time, 2) constant correlation across same-subject
measurements, and 3) observations from each subject are obtained at all
time points through the study (a condition also known as \emph{complete
observations}) \citep{gueorguieva2004} \citep{schober2018}.

The expected linear behavior of the response through time is a key
requisite in rm-ANOVA \citep{pinheiro2006}. This ``linearity
assumption'' in rm-ANOVA implies that the model is misspecified when the
data does not follow a linear trend, which results in unreliable
inference. In biomedical research, non-linear trends are the norm rather
than the exception in longitudinal studies. A particular example of this
non-linear behavior in longitudinal data arises in measurements of tumor
response to chemo and/or radiotherapy in preclinical and clinical
settings \citep{roblyer2011} \citep{skala2010} \citep{vishwanath2009}.
These studies have shown that the collected signal does not follow a
linear trend over time, and presents extreme variability at different
time points, making the fit of rm-ANOVA model inconsistent with the
observed variation. Therefore, when rm-ANOVA is used to draw inference
of such data the estimates are inevitably biased, because the model is
only able to accommodate linear trends that fail to adequately represent
the biological phenomenon of interest.

@@

As the rm-ANOVA and the more flexible LMEM approaches make overly
restrictive assumptions regarding the linearity of the response, there
is a need for biomedical researchers to explore the use of additional
statistical tools that allow the data (and not an assumption in trend)
to determine the trend of the fitted model, to enable appropriate
inference. In this regard, generalized additive models (GAMs) present an
alternative approach to analyze longitudinal data. Although not
frequently used by the biomedical community, these semi-parametric
models are customarily used in other fields to analyze longitudinal
data. Examples of the use of GAMs include the analysis of temporal
variations in geochemical and palaeoecological data
\citep{rose2012, pedersen2019, simpson2018}, health-environment
interactions \citep{yang2012} and the dynamics of government in
political science \citep{beck1998}. There are several advantages of GAMs
over LMEMs and rm-ANOVA models: 1) GAMs can fit a more flexible class of
smooth responses that enable the data to dictate the trend in the fit of
the model, 2) they can model non-constant correlation between repeated
measurements \citep{wood2017} and 3) can easily accommodate missing
observations. Therefore, GAMs can provide a more flexible statistical
approach to analyze non-linear biomedical longitudinal data than LMEMs
and rm-ANOVA.

@@

\bibliographystyle{sagev}
\bibliography{refs.bib}
